Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-15 src/ms.tex
addressing co-author comments

Unlike the long-period pile-up, the short-period pile-up is not predicted by the WMB model. However, this feature may also be due to an epoch of stalled wind-driven angular momentum losses, albeit a temporary one as cluster studies demonstrate that stars continue to spin down beyond the short-period pile-up. Works examining the \teff–\prot sequences of open clusters have found overlap between low-mass members in clusters of different ages, notably Praesepe (0.67~Gyr), NGC 752 (1~Gyr), and NGC 6811 (1.4~Gyr) \citep{Agueros2018, Curtis2019a, Curtis2020}. In other words, the spin rates of low-mass stars appear to evolve very little in the time elapsed between the ages of those clusters. The short-period pile-up we observe may be the manifestation of the same type of slowed spin evolution, but for stars of higher masses and younger ages (since the pile-up more closely aligns with the Pleiades sequence) than the cluster members in the above-mentioned works.

\citet{Curtis2019a, Curtis2020} proposed that such an effect could be produced by a temporary epoch of stalled spin-down, caused either by (a)a reduction in the magnetic braking torque, or (b) core–envelope momentum transfer which offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}. In the core–envelope momentum transfer scenario angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. Theoretical predictions for the core–envelope coupling are mass-dependent, and for a solar-mass star range from 30--110 Myr \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}. In the core–envelope coupling scenario, wind-driven spin-down stalls when the star's age becomes comparable to the core–envelope coupling timescale. Stars then resume spinning down via magnetic braking, and from open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. This timescale is consistent with independent studies of open clusters, which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

    \item The existence of the long-period pile-up limits the utility of gyrochronology for the hottest stars with convective envelopes, as stellar spin-down appears to stall on the ridge. For example, a Sun-like star may spend several Gyr evolving across the ridge. Authors using rotation periods as means of identifying new members to coeval stellar associations should be cognizant of this pile-up for stars with \teff~$\gtrsim5800~K$ as it can mimic the the appearance of a coeval stellar association in the \teff--\prot plane.

    \item An increasing number of open clusters are being discovered by searching for a clustering of rotation periods along a slow-rotator sequence. (I think maybe Pisces Eridani is an example of this? Or perhaps Jason has a better example?). However, the long-period pile up discovered here can mimic a slow-rotator sequence in a small sample of unassociated stars with different ages and precisely measured temperatures and rotation periods. This is clearly demonstrated by the CKS sample in the right-hand panel of figure 1 -- taken out of context, these data look like a group of coeval stars with a slow-rotator sequence. The discovery of pile-ups in stellar rotation periods therefore has consequences for open cluster studies. When over-densities or ridges appear present in the rotation period distributions of stellar populations, care must be taken to ensure that these are not caused by WMB or core-envelope coupling before assuming that population is coeval and using the any over-densities to age-date it.

>>>>

Unlike the long-period pile-up, the short-period pile-up is not predicted by the WMB model. However, this feature may also be due to an epoch of stalled wind-driven angular momentum losses, albeit a temporary one since cluster studies demonstrate that stars continue to spin down beyond the short-period pile-up. Works examining the \teff–\prot sequences of open clusters have found overlap between low-mass members in clusters of different ages, notably Praesepe (0.67~Gyr), NGC 752 (1~Gyr), and NGC 6811 (1.4~Gyr) \citep{Agueros2018, Curtis2019a, Curtis2020}. In other words, the spin rates of low-mass stars appear to evolve very little in the time elapsed between the ages of those clusters. The short-period pile-up we observe may be the manifestation of the same type of slowed spin evolution, but for stars of higher masses and younger ages than the cluster members in the above-mentioned works (since the pile-up is observed at shorter periods relative to the Praesepe sequence).

\citet{Curtis2019a, Curtis2020} proposed that the overlapping cluster sequences could be produced by a temporary epoch of stalled spin-down, caused either by (i) a reduction in the magnetic braking torque, or (ii) core–envelope momentum transfer which offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}. In the core–envelope momentum transfer scenario angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. The angular momentum transfer spins up the envelope, temporarily offsetting the spin-down via magnetic braking. Thus, in the core–envelope coupling scenario, spin evolution is slowed when the star's age becomes comparable to the core–envelope coupling timescale. Theoretical predictions for the core–envelope coupling are mass-dependent, and for a solar-mass star range from 30--110 Myr \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}.

After a period of slowed spin-evolution, stars must resume spinning down via magnetic braking as evidenced from studies of older open clusters. From open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. Thus, if core–envelope coupling is responsible for delaying the spin-down of stars, and if the theoretical core–envelope coupling timescales are accurate, then we may expect Sun-like stars to experience little spin evolution between $\sim$100~Myr and $\sim$200~Myr. This picture is compatible with open cluster studies which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$200--300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

    \item The existence of the long-period pile-up limits the utility of gyrochronology for the hottest stars with convective envelopes, as stellar spin-down appears to stall on the ridge. For example, a Sun-like star may spend several Gyr evolving across the ridge. \textbf{Authors using rotation periods as means of identifying new members to coeval stellar associations should be cognizant of this pile-up for stars with \teff~$\gtrsim5800~K$ as it can mimic the the appearance of a coeval stellar association in the \teff--\prot plane.}

    \item An increasing number of open clusters are being discovered by searching for a clustering of rotation periods along a slow-rotator sequence. However, the long-period pile up discovered here can mimic a slow-rotator sequence in a small sample of unassociated stars with different ages and precisely measured temperatures and rotation periods. This is clearly demonstrated by the CKS sample in the right-hand panel of Figure~\ref{fig:surveys}. Taken out of context, this sample resembles a group of coeval stars with a slow-rotator sequence. The discovery of pile-ups in stellar rotation periods therefore has consequences for open cluster studies. When an overdensity or ridge is present in the rotation period distribution of a stellar population, care must be taken to ensure that it is not caused by WMB or core-envelope coupling before assuming that population is coeval and using the overdensity to age-date it.
